---
title: "DYexperiments"
author: "Daisy Yu"
date: "11/8/2020"
output: pdf_document
---


## Case control simulation with population as a confounder
```{r}
library(dplyr)
library(ggplot2)

simSNPCov_popl=function(n,Beta=log(rf(K,m,m)),MAF1,MAF2,beta_s){
  ncase <- ncon <- n/2
  conX <- matrix(NA,ncol=K+1,nrow=ncon)
  caseX <- matrix(NA,ncol=K+1,nrow=ncase)
  
  # Simulate population status first (population is the K+1 column)
  f_0 <- f_1 <- 0.5
  conX[,K+1] <- rbinom(ncon,1,f_0)
  caseX[,K+1] <- sample(0:1,size=ncase,replace=TRUE,prob=c(f_0,f_1*exp(beta_s)))
  
  # Simulate SNVs conditional on population status
  for (i in 1:K){ 
    maf0 <- MAF1[i]
    maf1 <- MAF2[i]
    beta <- Beta[i]
    
    ## control
    con_pop0 <- which(conX[,K+1]==0)
    con_pop1 <- which(conX[,K+1]==1)
    
    conX[con_pop0,i] <- rbinom(length(con_pop0),size=2,prob=maf0)
    conX[con_pop1,i] <- rbinom(length(con_pop1),size=2,prob=maf1)
    
    ## case
    case_pop0 <- which(caseX[,K+1]==0)
    case_pop1 <- which(caseX[,K+1]==1)
    
    pp0 <- c((1-maf0)^2,2*maf0*(1-maf0)*exp(beta),maf0^2*exp(2*beta))
    pp1 <- c((1-maf1)^2,2*maf1*(1-maf1)*exp(beta),maf1^2*exp(2*beta))
    caseX[case_pop0,i] <- sample(0:2,size=length(case_pop0),replace=TRUE,prob=pp0)
    caseX[case_pop1,i] <- sample(0:2,size=length(case_pop1),replace=TRUE,prob=pp1)
  }
  X <- rbind(caseX,conX)
  colnames(X) <- c(paste0("X",1:K),"Population");rownames(X) <- NULL
  case <- c(rep(1,ncase),rep(0,ncon))
  return (data.frame(X,case))
}
```


### Sample MAF from the joint SFS of CEU and YRI populations.
```{r}
set.seed(6666)
load("~/Desktop/Simulation/1000GenomesProject/Input_dataset_1000GP_CEU.RData")
load("~/Desktop/Simulation/1000GenomesProject/Input_dataset_1000GP_YRI.RData")
m=4
K=50
index <- intersect(colnames(X_CEU),colnames(X_YRI))
X_CEU <- X_CEU[,index]
X_YRI <- X_YRI[,index]
MAF_CEU <- colSums(X_CEU)/(2*dim(X_CEU)[1])
MAF_YRI <- colSums(X_YRI)/(2*dim(X_YRI)[1])
Snp_data <- data.frame(CEU=MAF_CEU,YRI=MAF_YRI)
Snp_data <- Snp_data[-which(Snp_data$CEU<=0.05 & Snp_data$YRI<=0.05),]
maf <- Snp_data[sample(nrow(Snp_data),K,replace=F),]
Beta <- log(rf(K,m,m))
sim_data <- simSNPCov_popl(200,Beta,MAF1 = maf[,1],MAF2 = maf[,2],beta_s=1)

betahat <- c()  # with popl
betahat0 <- c() # without popl
for (i in 1:K) {
  XX <- sim_data %>% dplyr::select(c(paste0("X",i),Population,case)) %>% dplyr::rename(X=paste0("X",i))
  betahat[i] <- coef(glm(case~.,data=XX,family=binomial(),maxit=100))[[2]]
  
  XX0 <- sim_data %>% dplyr::select(c(paste0("X",i),case)) %>% dplyr::rename(X=paste0("X",i))
  betahat0[i] <- coef(glm(case~.,data=XX0,family=binomial(),maxit=100))[[2]]
}

Confounder=c()
for (i in 1:K){
  if (abs(betahat[i]-betahat0[i])/abs(betahat[i]) > 0.1) {Confounder[i]=T}
  else {Confounder[i]=F}
}
sum(Confounder)
maf <- cbind(maf,Confounder)

X1 <- data.frame(value=betahat)
X2 <- data.frame(value=betahat0)
X1$var <- "with_popl"
X2$var <- "without_popl"
data <- rbind(X1, X2)
ggplot(data, aes(value, fill = var)) + geom_density(alpha = 0.2) + xlim(c(-5,5))
```
23 out of 50 coefficients are changed by more than 10\% when population is excluded. The distribution of estimated coefficients doesn't change a lot with or without population included in the model, so we should get a similar $m$. 


### What if we only restrict to SNPs whose MAF differs by more than 0.3 between the two populations ? 
```{r}
set.seed(6666)
Snp_data <- Snp_data[which(abs(Snp_data$CEU-Snp_data$YRI) >= 0.3),]
maf <- Snp_data[sample(nrow(Snp_data),K,replace=F),]
Beta <- log(rf(K,m,m))
sim_data <- simSNPCov_popl(200,Beta,MAF1=maf[,1],MAF2=maf[,2],beta_s=1)

betahat <- c()  # with popl
betahat0 <- c() # without popl
for (i in 1:K) {
  XX <- sim_data %>% dplyr::select(c(paste0("X",i),Population,case)) %>% dplyr::rename(X=paste0("X",i))
  betahat[i] <- coef(glm(case~.,data=XX,family=binomial(),maxit=100))[[2]]
  
  XX0 <- sim_data %>% dplyr::select(c(paste0("X",i),case)) %>% dplyr::rename(X=paste0("X",i))
  betahat0[i] <- coef(glm(case~.,data=XX0,family=binomial(),maxit=100))[[2]]
}

Confounder=c()
for (i in 1:K){
  if (abs(betahat[i]-betahat0[i])/abs(betahat[i]) > 0.1) {Confounder[i]=T}
  else {Confounder[i]=F}
}
sum(Confounder)
maf <- cbind(maf,Confounder)

X1 <- data.frame(value=betahat)
X2 <- data.frame(value=betahat0)
X1$var <- "with_popl"
X2$var <- "without_popl"
data <- rbind(X1, X2)
ggplot(data, aes(value, fill = var)) + geom_density(alpha = 0.2) + xlim(c(-5,5))
```
Now 43 out of 50 coefficients are changed by more than 10\% when population is excluded. The distribution of estimated coefficients with population included in the model has a heavier tail than the distribution without population (i.e. coefficients are more extreme when we include population as a confounder). In this case, I guess we should get a much smaller $m$ when we include population. 
