---
title: "Data simulation experiments"
author: "BM"
date: "23/10/2020"
output: pdf_document
---

## Case control simulation 1

1. Sample covariates on a population,
2. sample covariate effects from log-F prior, 
3. calculate linear predictor of the logistic regression 
using all covariate effects,
4. simulate disease status, and
5. sample cases and controls from population.

```{r}
ncase <- ncon <- 100
K <- 50 # num SNPs
m <- 4 # log-F parameter
N <- 20000 # popn size
beta0 <- (-4) # intercept

ccsim1 <- function(MAF) {
  X <- cbind(1,matrix(rbinom(N*K,size=2,p=MAF),ncol=K))
  beta <- c(beta0,log(rf(K,m,m)))
  linpred <- X %*% beta
  p <- exp(linpred)/(1+exp(linpred))
  cc <- rbinom(N,size=1,prob=p)
  caseind <- sample((1:N)[cc==1],size=ncase,replace=FALSE)
  conind <- sample((1:N)[cc==0],size=ncon,replace=FALSE)
  list(cc=c(rep(1,ncase),rep(0,ncase)),X = X[c(caseind,conind),],
       beta = beta)
}
```

* Start with simulations for MAF = 0.5

```{r}
set.seed(123)
dat <- ccsim1(MAF=.5)
betahat <- rep(NA,K)
for(i in 1:K) {
  XX <- dat$X[,1+i]
  betahat[i] <- coef(glm(dat$cc~XX,family=binomial()))[[2]]
}
plot(dat$beta[-1],betahat) # correlated, but betahats attenuated 
hist(betahat,freq=FALSE,nclass=10,ylim=c(0,.8),xlim=c(-3,3))
hist(dat$beta[-1],freq=FALSE,nclass=10,ylim=c(0,.8),xlim=c(-3,3))
```

* $\hat{\beta}$'s are correlated with the $\beta$'s but are 
attenuated. I guess this is why DY and SC's methods are suggesting
that $m$ is very large (variance of log-F small).

* Next try MAF = 0.5

```{r}
set.seed(123)
dat <- ccsim1(MAF=.05)
betahat <- rep(NA,K)
for(i in 1:K) {
  XX <- dat$X[,1+i]
  betahat[i] <- coef(glm(dat$cc~XX,family=binomial()))[[2]]
}
plot(dat$beta[-1],betahat) # betahats still attenuated, but less so 
hist(betahat,freq=FALSE,nclass=10,ylim=c(0,.8),xlim=c(-3,3))
hist(dat$beta[-1],freq=FALSE,nclass=10,ylim=c(0,.8),xlim=c(-3,3))
```

* $\hat{\beta}$'s are less attenuated, but still too small in magnitude.
* Consistent with the fact that DY and SC's likelihoods for m sometimes
have a max.

* Can we fit a joint model in X's, like the one used to 
simulate the data? No, it fails with current sample size.

```{r}
XX <- dat$X[,-1]
colnames(XX)<- paste0("X",1:K)
dd <- data.frame(cc=dat$cc,XX)
gg <- glm(cc ~ ., data=dd, family=binomial())
head(round(cbind(dat$beta,coef(gg),c(1,betahat)),3)) # too unstable
plot(dat$beta[-1],coef(gg)[-1])
```

## Case control simulation 2

* Use the Qin and Zhang method.
* If $g(x)$ is the covariate distribution in controls, 
the distribution in cases is
proportional to $g(x) \exp(x \beta)$. 
* If $g(x)$ is $binomial(2,p)$, where $p$ is the MAF (under
a rare disease, the distribution in controls is about that 
in the population), then

$$
g(x) \exp(x \beta) = \left\{ 
\begin{array}{rl}
(1-p)^2 & x=0 \\
2p(1-p) \exp(\beta) & x = 1 \\
p^2 \exp(2\beta) & x = 2 \end{array}
\right. ,
$$

which has normalizing constant 
$(1-p)^2 + 2p(1-p) \exp(\beta) + p^2 \exp(2\beta)$.

```{r}
rcase <- function(n,beta,MAF) {
  p0 <- (1-MAF)^2
  p1 <- 2*MAF*(1-MAF)*exp(beta)
  p2 <- MAF^2*exp(2*beta)
  pp <- c(p0,p1,p2)
  sample(0:2,size=n,replace=TRUE,prob=pp)
}

ccsim2 <- function(MAF) {
  Xcon <- matrix(rbinom(ncon*K,size=2,prob=MAF),ncol=K)
  beta <- log(rf(K,m,m))
  Xcase <- matrix(NA,ncol=K,nrow=ncase)
  for(i in 1:K) {
    Xcase[,i] <- rcase(ncase,beta[i],MAF)
  }
  X <- rbind(Xcase,Xcon)
  list(cc=c(rep(1,ncase),rep(0,ncase)),X = X,beta = beta)
}
```

* Simulate covariates with MAF = 0.05 and fit single-SNP models
to see what the coefficients look like.
    * We see one estimated coefficient of about $-15$, which I guess
    means non-convergence or monotone likelihood for that covariate.


```{r}
set.seed(123)
dat <- ccsim2(MAF=.05)
betahat <- rep(NA,K)
for(i in 1:K) {
  XX <- dat$X[,i]
  betahat[i] <- coef(glm(dat$cc~XX,family=binomial()))[[2]]
}
plot(dat$beta,betahat) 
hist(betahat,freq=FALSE,nclass=30,ylim=c(0,.4),xlim=c(-3,3))
hist(dat$beta,freq=FALSE,nclass=30,ylim=c(0,.4),xlim=c(-3,3))
```



* Repeat with MAF = 0.5

```{r}
set.seed(123)
dat <- ccsim2(MAF=.5)
betahat <- rep(NA,K)
for(i in 1:K) {
  XX <- dat$X[,i]
  betahat[i] <- coef(glm(dat$cc~XX,family=binomial()))[[2]]
}
plot(dat$beta,betahat) 
hist(betahat,freq=FALSE,nclass=30,ylim=c(0,.4),xlim=c(-3,3))
hist(dat$beta,freq=FALSE,nclass=30,ylim=c(0,.4),xlim=c(-3,3))
```

